---
title: Section 01 Introduction
subtitle: "Gov 1347: Election Analytics"
author: Kiara Hernandez
date: \today
institute: Harvard University
fontsize: 20pt
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
# set working directory always
knitr::opts_knit$set(root.dir = "~/Desktop/GOV1347.nosync")
#options(repos = list(CRAN="http://cran.rstudio.com/"))
# required packages 
require(tidyverse)
require(ggplot2)
require(sf)
```

# Interactive Session in \texttt{R Studio}

## Geographic and temporal trends in midterm elections?

```{r part1, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE, fig.align="center"}
## make map of vote share by state and CD

# start with 114th congress - 2014 election

# load geographic data
get_congress_map <- function(cong=114) {
  tmp_file <- tempfile()
  tmp_dir  <- tempdir()
  zp <- sprintf("https://cdmaps.polisci.ucla.edu/shp/districts114.zip",cong)
  download.file(zp, tmp_file)
  unzip(zipfile = tmp_file, exdir = tmp_dir)
  fpath <- paste(tmp_dir, sprintf("districtShapes/districts114.shp",cong), sep = "/")
  st_read(fpath)
}

# load 114th congress
cd114 <- get_congress_map(114)

# select specific state 
cd114_nj <- cd114 %>% 
            filter(STATENAME=="New Jersey") %>%
            mutate(DISTRICT = as.character(DISTRICT))%>%
            select(DISTRICT)
```

```{r part2, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE, fig.align="center"}
# add data to plot - 2014 GOP party seat share
# reload election data - h from previous exercise
h <- read_csv("data/house party vote share by district 1948-2020.csv")

# filter for 2014 election and state
R_nj_2014 <- h %>%
    filter(raceYear == 2014, State == "New Jersey") %>%
    select(raceYear, State, district_num, RepVotesMajorPercent, DemVotesMajorPercent) %>%
  # summarize party vote share by district
    group_by(district_num) %>%
    summarise(Rep_votes_pct = RepVotesMajorPercent) %>%
  # rename district variable name to match shapefile
    rename(DISTRICT = district_num)

# before joining dfs, check classes of variable to be merged on
class(R_nj_2014$DISTRICT)
class(cd114_nj$DISTRICT)

# change class
cd114_nj$DISTRICT <- as.numeric(cd114_nj$DISTRICT)

# join election returns with shapefiles
cd114_nj <- cd114_nj %>% left_join(R_nj_2014, by="DISTRICT")
cd114_nj
```

```{r part3, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE, fig.align="center"}
# time to map!
cd114_nj %>% ggplot() + 
  geom_sf(aes(fill=Rep_votes_pct),
          inherit.aes=FALSE,alpha=0.9) + 
  scale_fill_gradient(low = "white", high = "black", limits=c(10,80)) +
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) 
```

```{r part4, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE, fig.align="center"}
## make map of GOP vote share by state (national) - 2014
# use h dataset from earlier
# house_party_vote_share_by_district_1948_2020 <- 
  # read_csv("house party vote share by district 1948-2020.csv")
# h <- house_party_vote_share_by_district_1948_2020
 
# filter for relevant variables
R_2014 <- h %>%
    filter(raceYear == 2014) %>%
    select(raceYear, State, district_num, district_id, RepVotes, DemVotes) %>%
  # summarize party vote share by state
    group_by(State) %>%
  # mutate Rep vote margin by state %>%
    mutate(R_votemargin_st = (sum(RepVotes))/
                            sum(RepVotes + DemVotes),
           D_votemargin_st = (sum(DemVotes))/
                            sum(RepVotes + DemVotes)) %>%
    rename(state = State)

# load usmap
# install.packages('plot_usmap')
library(usmap)

states_map <- usmap::us_map()
unique(states_map$abbr)

# plot
plot_usmap(data = R_2014, regions = "states", values = "R_votemargin_st") + 
  scale_fill_gradient(low = "white", high = "red", name = "GOP two-party voteshare margin") +
  theme_void()
```

```{r part5, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE, fig.align="center"}
## map across all election years
# filter for relevant variables
R_all <- h %>%
    select(raceYear, State, district_num, district_id, RepVotes, DemVotes) %>%
  # summarize party vote share by state
    group_by(State) %>%
  # mutate Rep vote margin by state %>%
    mutate(R_votemargin_st = (sum(RepVotes))/
                            sum(RepVotes + DemVotes),
           D_votemargin_st = (sum(DemVotes))/
                            sum(RepVotes + DemVotes)) %>%
    rename(state = State)

# plot
plot_usmap(data = R_all, regions = "states", values = "R_votemargin_st") + 
  facet_wrap(facets = raceYear ~.) +
  scale_fill_gradient(low = "white", high = "red", name = "GOP two-party voteshare margin") +
  theme_void()

```

A summary of the district-level exercise (and lots of other helpful information about
mapping congressional districts in R) can be found here:
<https://cdmaps.polisci.ucla.edu/tut/mapping_congress_in_R.html>

# Final thoughts

-   Impossible to memorize every single `dplyr` and `ggplot2` command
    you'll need. 
-   Recommendation: create R Studio snippets for various plots we've
    shown today (`Preferences > Code > Snippets`).
-   Visualization is about communication, not coding ... points
    for **clarity** and **engagement**, not technical
    wizardry.
-   Lots of `ggplot2` tools to creatively communicate:
    \url{exts.ggplot2.tidyverse.org/gallery/}

# Blog extensions

1.  **Visualization customization.** Use the tools from this
    section to 
    1. create a map of Republican/Democrat voteshare margin
    by state in a year of your choice,
    2. create a map of Republican/Democrat voteshare margin by state
         and congressional district in 2014,
    3. label each state (e.g. `AZ`) in your map and
    4. create a custom `ggplot2` theme for your blog.
    
2.  **Gerrymandering extension.** So far, we've been looking at the
    voteshare margin by party. This means we have been working with the
    popular vote as opposed to the number of seats won in the House
    (recall our discussion from Tuesday about the inefficient geographic
    distribution of Democrats as compared to Republicans that results
    from gerrymandering --- the drawing of district boundaries to give
    parties electoral advantages. Read more here:
    \url{https://www.washingtonpost.com/news/wonk/wp/2015/03/01/this-is-the-best-explanation-of-gerrymandering-you-will-ever-see/},
    Why Cities Lose: The Deep Roots of the Urban-Rural Political Divide.
    Jonathan A. Rodden. Basic Books. 2019.).
    * Create a map of seat share by party for the entire U.S. in a year of
    your choice. 
      + You can find and download the data at this website:
      \url{https://guides.library.harvard.edu/hks/campaigns_elections>}. 
      + Select "CQ Voting & Elections Collection" \> Election Results \> Office:
      + House Elections \> Election Type: General \> Region: National \>
      + Year: [your choice]. 
      + Similar to the map you created of voteshare
      margin by state, your map should contain all 50 states shaded
      according the proportion of seats won by Republicans/Democrats. How do
      seat share and voteshare compare?
    
3.  \textbf{Swing state map extension.} Instead of plotting
    \textit{voteshare-margin} maps for each year $y$ where the color for
    a state is colored by the quantity $\frac{R_y}{D_y+R_y}$ (or
    $\frac{D_y}{D_y+R_y}$), plot a \textit{swing map} for each year
    where each state is colored by the quantity
    $\frac{R_y}{D_y+R_y} - \frac{R_{y-4}}{D_{y-4} + R_{y-4}}$. Which
    states are/have been battleground states? Which are no longer?

